---
title: "Probability"
author: "Dave Tang"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(
  library(tidyverse)
)
```

A lot of bioinformatics tools use probabilistic models, so it's important to
have at least a working understanding of probability. In simpler cases, if you
can figure out **all the possible events**, you can calculate the probability
of a subset of events by dividing the subset by all possible events. For
example, in a full deck of cards, there are 52 cards and the probability of
picking a red card is the subset of red cards (26) divided by all the cards
(52).

Another way to think about the probability of an event is that it is the
proportion of times that the event occurs when we repeat the experiment an
infinite number of times, independently and under the same conditions. The
probability is the number of times an event occurred divided by the total
number of times an experiment was repeated. Because of this, we can get a
estimate of the probability of an event by running Monte Carlo simulations.
This is particularly useful when it is difficult to work out the probability
mathematically.

## Basics

* $Pr(A)$ is the probability of event $A$ happening; the general term _event_
  is used to refer to things that can happen by chance.

* A probability distribution is the probabilities of all possible events.

* For discrete variables, the probability distribution is calculated using a
  probability density function (PDF).

* For continuous variables, the probability distribution is calculated using
  the cumulative distribution function (CDF).

* Two events are independent if the outcome of one does not affect the other.
  The classic example is coin tosses; every time we toss a fair coin, the
  probability of seeing heads is 1/2 regardless of what previous events were.

When events are not independent, conditional probabilities are used.

$$ Pr(Card\ 2\ is\ a\ king\ |\ Card\ 1\ is\ a\ king) = \frac{3}{51} $$

The $|$ means "given that" or "conditional on". The mathematical definition of independence is:

$$ Pr(A\ |\ B) = Pr(A) $$

The fact that event $B$ has occurred does not affect the probability of $A$.

We use the multiplication rule for calculating the probability of two events occurring:

$$ Pr(A\ and\ B) = Pr(A) \cdot pr(B\ |\ A) $$

When there are three events:

$$ Pr(A\ and\ B\ and\ C) = Pr(A) \cdot Pr(B\ |\ A) \cdot Pr(C\ |\ A\ and\ B) $$

If the events are independent, then:

$$ Pr(A\ and\ B\ and\ C) = Pr(A) \cdot Pr(B) \cdot Pr(C) $$

General formula for computing conditional probabilities:

$$ Pr(B\ |\ A) = \frac{Pr(A\ and\ B)}{Pr(A)} $$

If event $A$ and $B$ are independent, then the probability just becomes $Pr(B)$:

$$ Pr(B\ |\ A) = \frac{Pr(A) \cdot Pr(B)}{Pr(A)} = Pr(B) $$

Addition rule for calculating the probability of either event happening (but not both, which is why we subtract):

$$ Pr(A\ or\ B) = Pr(A) + Pr(B) - Pr(A\ and\ B) $$

Bayes' theorem:

$$ Pr(A\ |\ B) = \frac{Pr(B\ |\ A) \cdot Pr(A)}{Pr(B)} $$

Again, if event $A$ and $B$ are independent, then the probability just becomes $Pr(B)$:

$$ Pr(A) = \frac{Pr(B) \cdot Pr(A)}{Pr(B)} $$

One application of Bayes' Theorem is for spam filtering where:

* Event A: the message is spam
* Text: the message contains certain words

$$ Pr(spam\ |\ words) = \frac{Pr(words\ |\ spam) \cdot Pr(spam)}{Pr(words)} $$

## Marginal and Conditional Probability

A 2 by 2 contingency table can help illustrate marginal and conditional probability.

```{r two_by_two_table}
disease_table <- matrix(
  c(40, 10, 20, 30),
  nrow = 2,
  dimnames = list(
    "Test" = c("Test +", "Test -"),
    "Disease" = c("Disease +", "Disease -")
  )
)

disease_table_with_margins <- addmargins(disease_table)

disease_table_with_margins
```

Marginal probabilities are read directly from the `totals` in the margins of the table, ignoring the other variable. Reading the `Sum`, probability of disease (`Disease +`) is 0.5 and probability to test positive (`Test +`) is 0.6.

```{r prop_table}
disease_table |>
  prop.table() |>
  addmargins()
```

Conditional probabilities restrict attention to a single row or column. Given a positive test, we restrict ourselves to the first row.

$P(Disease+|Test+)$ is of those who tested positive, what fraction have disease?

$$
P(Disease+|Test+) = 0.4 / 0.6 = 0.67
$$

Given positive to disease, we restrict ourselves to the first column.

$P(Test+|Disease+)$ is of those with disease, what fraction test positive?

$$
P(Test+|Disease+) = 0.4 / 0.5 = 0.80
$$

Note that $P(Disease+|Test+)$ is not equal to $P(Test+|Disease+)$; conditional probability is not symmetric. This asymmetry is the source of the base rate fallacy, which is commonly encountered
in medical testing and legal reasoning, where $P(Disease|Test+)$ is confused
with $P(Test+|Disease)$.

## Bayes' theorem

A cell in the two by two contingency table is simply a tally, a count of people who satisfy two conditions simultaneously. The top-left cell is how many people in our total of 100 had both a positive test and the disease.

There are two different ways to express the probability of landing in any single cell of the table.

```{r disease_table_with_margins}
disease_table |>
  addmargins()
```

Consider the top-left cell with 40 people who are both Test+ and Disease+.

1. Start with Disease, then look at Test.

If we pull aside everyone with disease, we end up with 50 people. Then from those 50, if we ask how many tested positive we will get 40.

* First taking a 50% slice of 100 (Disease+)
* Then taking an 80% slice of those (ones who also tested positive)

$$
50\%×80\%=40\%
$$

2. Start with Test, then look at Disease

Again, if we pull aside everyone who tested positive, we end up with 60 people. Then from those 60, if we ask how many actually have the disease we get 40 again.

* First taking a 60% slice of 100 (Test+)
* Then taking a 67% slice of those (the ones who also have disease)

$$
60\%×67\%=40\%
$$

1. is actually the same as

$$
P(Disease+) \times P(Test+∣Disease+)
$$

* $P(Disease+)=50/100=50\%$, which is pulling aside Disease+ (marginal)
* $P(Test+∣Disease+)=40/50=80\%$, which is then asking among Disease+, how many tested positive (conditional)

2. is the same as

$$
P(Test+) \times P(Disease+∣Test+)
$$

* $P(Test+)=60/100=60\%$, which is pulling aside Test+ (marginal)
* $P(Disease+∣Test+)=40/60=67\%$, which is then asking among Test+, how many have disease (conditional)

Therefore:

$$
P(Disease+) \times P(Test+∣Disease+) = P(Test+) \times P(Disease+∣Test+)
$$

So if we want to know $P(Disease+ | Test+)$ but only have $P(Test+ | Disease+)$,
we simply rearrange:

$$
P(Test+) \times P(Disease+∣Test+)  = P(Disease+) \times P(Test+∣Disease+) \\

P(Disease+∣Test+) = \frac{P(Disease+) \times P(Test+∣Disease+)}{P(Test+)}
$$

This is Bayes' theorem. In the abstract form with A and B:

$$
P(A | B) = \frac{P(B | A) \times P(A)}{P(B)}
$$
